|

### 2021-06-15 04-Chp-CPRD-Analysis.Rmd
End of day: -2722

```{r cprdFlowchart, echo = FALSE, results="asis", fig.pos = "H", fig.cap='(ref:cprdFlowchart-cap)', out.width='100%', fig.scap='(ref:cprdFlowchart-scap)'}
knitr::include_graphics(file.path("figures","cprd-analysis","cohort_attrition_test.png"))
```
<!--------------------------------------------------------------------------->

@@

Full covariate information was available for `r missingtext`. Six key variables had some missing data: IMD 2010 score was missing for `r imdtext`, because it is only recorded for English practices; alcohol status was missing for `r alcoholtext`; smoking status was missing for `r smokingtext`; BMI, or a calculated BMI from height and weight measurements, was missing for `r bmitext`; baseline total cholesterol was missing for `r choltext`, and baseline LDL cholesterol was missing for `r ldltext`

>>>>

```{r cprdFlowchart, echo = FALSE, results="asis", fig.pos = "H", fig.cap='(ref:cprdFlowchart-cap)', out.width='100%', fig.scap='(ref:cprdFlowchart-scap)'}
knitr::include_graphics(file.path("figures","cprd-analysis","cohort_attrition.png"))
```
<!--------------------------------------------------------------------------->

@@

Full covariate information was available for `r missingtext`. Six key variables had some missing data: IMD 2010 score was missing for `r imdtext`, because it is only recorded for English practices; alcohol status was missing for `r alcoholtext`; smoking status was missing for `r smokingtext`; BMI, or a calculated BMI from height and weight measurements, was missing for `r bmitext`; baseline total cholesterol was missing for `r choltext`; and baseline LDL cholesterol was missing for `r ldltext`.

### 2021-06-15 manuscripts/cprd/manuscript.Rmd
End of day: -2722

---
bibliography: bibliography/references.bib
csl: bibliography/nature.csl
output:
  bookdown::pdf_document2:
    template: templates/brief_template.tex
  bookdown::word_document2: 
      toc: false
      toc_depth: 3
      reference_docx: templates/word-styles-reference-01.docx
      number_sections: false 
  bookdown::html_document2: default
documentclass: book
---

```{block type='savequote', include=knitr::is_latex_output(), quote_author='(ref:cprdquote)', echo = FALSE}
When dealing with human beings controlled experiments frequently prove to be impracticable, so for a scientific basis for our assumptions we turn to past history to reconstruct the suspected causal chain of events - and then our statistical troubles may begin.
```
(ref:cprdquote) --- Harold F. Dorn, 1953 [@dorn1953]

# Primary analysis of lipid regulating agents and dementia {#cprd-analysis-heading}

\minitoc <!-- this will include a mini table of contents-->

<!-- Include the early analysis we did here, that did not work as a result of limited scope. Could include some of the genetic work also. Will additionally need to put the relevant materials from the paper in the  -->

```{r, echo = FALSE, warning=FALSE, message=FALSE}
source("R/doc_options.R")
source("R/helper.R")
knitr::read_chunk("R/Chp3-CPRD-Analysis.R")
doc_type <- knitr::opts_knit$get('rmarkdown.pandoc.to') # Info on knitting format
```

@@

## Additional ideas

One of the problems I'll point out with the trials in the systematic review chapter (Chapter \@ref(sys-rev-heading)) is that they only included people with a high cardiovascular risk - we are kind of doing the same by using elevated test cholesterol results as the index event.

Might be able to get around this by pointing out that you need some index event to be able to do this analysis (?)

__Things we tried__

- Conditioning entry to the cohort on QRISK score (as defined by codes) rather than by . Need to be able to demonstrate here that not many meeting our original entry criteria actually go on to have a statin quickly. Need total number of those starting and average time to start.
- Positive controls. Need to be able to explain why these controls have such wildly increased results. Cross-reference these results with the Hernan paper.
- Competing risk analysis, with death as a competing risk. Need numbers of deaths in each group, and preliminary results
- Allowing for time-varying confounders for binary covariates.

__Things we could try__

- Work through data using positive control to see if the way it is being added is what is causing the issues.
- Examine Cramer et al. thesis and see what they did
- Marginal structural models approach (a la J. Sterne)

## Lay summary
Large scale database such as this are often called population level, and studies that make use of them have a number of advantages.

This analysis uses one of these databases, the Clinical Practice Research Datalink (CPRD), which contains the electronic medical records of XXXXXX people from YYYYYY general practices across England.

Using this data, I have analysed the ...

## Introduction

Lack of evidence for the association of statins on vascular dementia risk

Large scale databases advantages over other means

Cross reference here as well re: the lack of information on Vascular dementia and the need for analyses at different sources of bias.

## Methods

### Data source {#cprd-data-source}

The Clinical Practice Research Datalink (CPRD) is a large, population-based, electronic health record database.[@herrett2015] It contains the primary care records for

Originally termed the General Practice Research Database,

The

Data completeness and collection of certain disease are heavily influenced by exercises that reward doctors for capturing certain diagnoses, such as the Quality and Outcomes Framework. However, these drives can oftern have unintended consequences [@liu2019]

The CPRD uses a coding system called Read code <!--- check capitalisation --->

### Study design and protocol

@@

<!-- Note: If protocol was written by Venexia, how do I reference this? -->

@@

### Study cohort

Participants were included in our study cohort if their record contained any of the following index events: a Read code for a diagnosis of hypercholesterolemia or related condition; a Read code for prescription of a lipid-regulating agent (such as statins); a total cholesterol test result of >4 mmol/L; or an LDL-c test result of >2 mmol/L.

@@

### Exposures

@@

### Outcomes

<!--- Include short bit on follow-up here, describing how and why it was defined --->

We considered five outcomes as part of this analysis: probable  Alzheimer’s disease, possible Alzheimer’s disease, vascular dementia, other dementia, and a composite all-cause dementia outcome (Supplementary Figure 1). When two or more outcomes were coded in a participant’s clinical record, a decision tree was used to differentiate between them (Supplementary Figure 1). The diagnosis date of the outcome was determined by the first record of a relevant code.

Participants were followed up, and were censored at the earliest of date of diagnosis, death, date of transfer out of the study cohort or end of follow-up ( <!--- Get date here ---> )

_As a positive control outcome, we included cornonary artery disease. Statins have a known effect on this outcome on the basis of randomised controlled trials._

We also used a negative control of backpain to investigate the potential for residual confounding.

@@

### Covariates

A range of additional variables were included in the analysis, intended to address differential distributions of potnetial confounding variables between those who had a

<!--- Need a lot more here - see other thesis for example of how covariates are described --->

The analysis was adjusted for a range of baseline covariates including sex,

grouped year of entry into the cohort (<2000, 2000-2004, 2005-2009, >2010),

Charlson co-morbidity index,

Index of Multiple Deprivation (IMD),

consultation rate,

alcohol (current, former,  never),

smoking (current, former,  never),

BMI,

baseline total cholesterol,

and history of cardiovascular disease, coronary bypass surgery, coronary artery disease, peripheral arterial disease, hypertension, chronic kidney disease, and Type 1 and Type 2 diabetes.

All covariates were determined at index and definitions for each can be found in the archived data repository reflecting this study <!--- link to Zenodo here --->

@@

### Missing data

<!--- Review materials from multiple imputation course --->

Missing data was handled using a multiple imputation approach. Variables with missing observations were identified, and

<!-- Binary variables were modelled using a -->

<!-- continuos -->

<!-- As per best practice, all other variables, including the out  -->

<!-- The MICE command in Stata was used -->

In the case of missing data, we used multiple imputation by chained equations (MICE) in STATA to create 20 imputed datasets. All covariates included in the analytic model were also included in the imputation model.

@@

### Time axis

<!-- In Cox models, there is the option to -->

<!-- Age vs time-since entry -->

<!-- Expand on use of age as the time scale, but describe the problems this introduced -->

To address the potential for immortal time bias, we employed a time-varying indicator of treatment status to correctly allocate time-at-risk to the exposed and unexposed groups.[@levesque2010] We ran two models, one unadjusted except for age, which was captured by using the participant's age as the time axis for the model,[@lamarca1998; @gail2009; @pencina2007] and a second adjusted for the additional covariates described above. Additional analyses stratified by outcome and drug class were also performed.

&nbsp;
<!----------------------------------------------------------------------------->

### Estimation methods

Potential biases included time varying confounding, selection bias due to censoring on death and 
We use a Cox proportional hazards model with a time-varying treatment indicator.

Expand on post estimation analyses, which showed that nothing except age had an effect.

$$h(t) = h_o(t) \times exp(b_1x_1 + b_2x_2 + ... +b_px_p)$$

where:

* $t$ is the survival time;
* $h(t)$ is the hazard function; and
* $x_1,x_2,...,x_p$ are the covariates which determine the hazard function, while $b_1,b_2,...,b_p$ are the coefficients for each covariate.
* $h_o(t)$ is the baseline hazard - when all $x_i$ are zero, the $exp()$ function resolves to 1.

As the values of $b_i$ increases, value of $exp(b_i)$ (i.e. the hazard ratio) increases in tandem, thus reducing the survival time, $t$.

__Bit on why Cox models are slightly different from others, due to proportional hazards assumptions__

Hazard ratios are

Test for non-proportionality are therefore very important, as models which fail

Problems with testing for non-proportional hazards when

<!----------------------------------------------------------------------------->
(ref:cprd-dag-1-cap) __Directed acyclic graph showing assumed relationship between exposure and covariate:__

(ref:cprd-dag-1-scap) Directed acyclic graph showing assumed relationship between exposure and covariate

```{r cprd-dag-1, echo = FALSE, results="asis", fig.cap='(ref:cprd-dag-1-cap)', out.width='100%', fig.scap='(ref:cprd-dag-1-scap)'}
knitr::include_graphics(file.path("figures/sys-rev/prisma_flow.png"))
```
<!----------------------------------------------------------------------------->

### Immortal time bias and time-varying treatment indicators

Immortal time bias is a type of misclassification bias. It occurs when

One potential way to address this is the use of a time-varying treatment indicator, a factor which modi

In the case of this analysis, this treatment indicator is statin use. Using codes for statins (see Appendix \@ref())

### Estimating the value of the time-varying confounders

Mean time from index event to first prescription of statins was 2.4 years. This negates the promised benefit of ruling out confounding by indication (where the test result leads to the prescription of the treatment and also increases the risk of the outcome, distorting the relationship between the two), as there is no relationship between index TC/LDL-c and eventual LRA prescription.

Additionally, the time between index event and prescription does lead to a problem in terms of time varying confounding, as an average time of 2.4 years between current measurement of the covariates and treatment switching means there is plenty of time for the value of the covariate to change. This is problematic when the decision to change treatments (in this case to move from no LRA use to LRA use) is influenced by a set of prognostic factors that in turn may have been influenced by the initial treatment decision, as is likely to be the case for a range of covariates included in the model. For example:

    _No CVD (t=0) -> No LRA (t=0) -> CVD (t=1) -> LRA (t=1) -> Dementia (t=2)_

In this case, the decision to move to LRA use is influenced by CVD status at _Time 1_, which will not be captured by adjusting only for CVD status at _Time 0_.

In practice, this means that the value of the prognostic factor should be regularly captured

However, in electronic health records, a change in the value of the prongostic factors is only important if it is recorded in a patients record, as for it to have an impact on treatment decisions, it must be recorded.

This means we can find the most recent value of the covariate before the switch and apply a marginal structural model approach, filling all values for that variable before the most recent measure with the baseline measurement, and all after the most recent meausure with the value of the most recent measure (on the basis that you won't go from having CVD back to not having CVD).

i.e.  
Timepoint 12345678  
CVD       00001111  
Treatment 00000111

Split into 3 month blocks since index event and use the same approach as above to work out the values of each covariate at each time point.

Note: this will be harder for things that are not dichotomous and can go up as well as down. Examples include total cholesterol and BMI, which can go up as well as down.

### The effect of total cholesterol or LDL-cholesterol on LRA prescription

It would be fair to assume that the baseline total cholesterol/LDL-cholesterol would at least in part predict the likelihood of someone being prescribed a statin.

However, this is not the case. Baseline cholesterol level are predicted to be a poorer instrument for than QRISK2 score,[@hippisley-cox2008] which estimates a patients' 10-year risk of a cardiovascular event. Current NICE guidelines state that those with a QRISK score of 10% or higher, and in whom lifestyle modification is ineffective/inappropriate, should received a lipid regulation agent. However, this analysis could not find any effect of QRISK2 scores on statins prescription levels at 6 months.

<!-- Need to cite Lauren's eventual paper here focusing on QRISK2, but also display a RD analysis of TC/LDL-c levels here on statins at 6 months. Need also to check, as Lauren mentioned she found some evidence that there is a relationship in practices that actually did what they should. -->

As expected, in a confirmatory analysis using lipid levels, there was no association between *the most recent total cholesterol or LDL-cholesterol reading in the CPRD and the treatment, indicating that adjusting for this variable was not required.*

<!--- This hasn't been done! Plus I do not think this is the case --->

### Replicating other analytical strategies

**Comparing and contrasting between different studies is particularly difficult** because of the impact that the use of different code list can have on the analysis [@wilkinson2018a; @mcguinness2019c] This highlights a particular challenge in comparing research across different time-periods and coding systems. Previous work has demonstrated that electronic health records have variable positive and negative predictive values, based substantially on the exact code-lists used to define each outcome.

In order to

As part of my exploration of the unexpected results, I attempted to replicate the analyses of another paper that used a different approach to

Having contacted the authors to address these concerns, it was noted that

Propensity scores are based on a range of baseline variables, and provide an estimates of the

They used propensity scores to attempt to matcch patients equally likely to recieve the vaccine

In addition, the need to contact the authors for their code-lists reinforces the points raised in Section \@ref(sys-rev-open-data) - that all relevant material should be readily shared. Fortunately, the authors kindly shared their

However, there was a large difference in the case mix of dementia diagnoses between the two analyses.

However, as shown in Table \@ref(tab:smeethComparison-table), there is substantial divergence between the two sets of code-lists.

The reason for the discrepancy in terms of the number of Alzheimer cases as a percentage of total dementia cases between our study and the Smeeth 2010 paper is due to two things:
•	While all of the codes used to define Alzheimer’s in the Smeeth paper are included in our “Probable Alzheimer’s” code-list, we also have several additional codes used to define this outcome (including, for example, “Eu00013: [X]Alzheimer's disease type 2”). 
•	Several of the codes in our “Possible Alzheimer’s” code list are included in the “Other dementia” code list used by Smeeth.

Using the Smeeth code-list, we get a similar breakdown of cases to that paper, with Alzheimer’s disease being a lot less frequent than when using our code-lists (which is not surprising based on the points above). <!-- This, in addition to unexplained statistical issues such as adjusting for post-baseline covariates, means I am sceptical of the value of comparing our findings with Smeeth. -->
IN addition, this analysis used propensity s. IN addition, an analytical approach which appeared to adjust for key variables likely to be on the analytical pathway between

Replicating a matched analysis a la Smeeth et al 2010 - though questions were raised as to how they had actually done the analysis (they did not match on propensity score, they simply adjusted for it - find paper that explores 4 ways of accounting for PS and shows this is the worst), they adjusted for things likely to be on the causal pathway, and they saw a huge change in direction for MI pre- vs post- adjustment.

@@

<!----------------------------------------------------------------------------->
(ref:smeethComparison-caption) Comparison of code lists used in this analysis and in Smeeth et al.

(ref:smeethComparison-scaption) Comparison of code lists used in this analysis and in Smeeth et al.

```{r smeethComparison-table, message=FALSE, results="asis", echo = FALSE}
```
<!----------------------------------------------------------------------------->

## Results

&nbsp;
<!----------------------------------------------------------------------------->

### Patient sample

<!----------------------------------------------------------------------------->
(ref:cprdFlowchart-cap) __Attrition of CPRD participants as the eligibility criteria were applied:__

(ref:cprdFlowchart-scap) Attrition of CPRD participants

```{r cprdFlowchart, echo = FALSE, results="asis", fig.pos = "H", fig.cap='(ref:cprdFlowchart-cap)', out.width='100%', fig.scap='(ref:cprdFlowchart-scap)'}
knitr::include_graphics(file.path("figures","cprd-analysis","cohort_attrition.png"))
```
<!----------------------------------------------------------------------------->

<!--- Add some text here about the fact highlighting the key criteria that excluded participants --->

### Patient characteristics

@@

<!--- Table 1 is missing type 2 diabetes --->

\blandscape
<!----------------------------------------------------------------------------->
(ref:cprdCharacteristics-caption) Patient characteristics by drug class

(ref:cprdCharacteristics-scaption) Patient characteristics by drug class

```{r cprdCharacteristics-table, message=FALSE, results="asis", echo = FALSE}
```
<!----------------------------------------------------------------------------->
\elandscape

@@

### Missing data

Full covariate information was available for `r missingtext`. Five key variables had some missing data: IMD 2010 score, a proxy for socioeconomic position that is measured as twentiles with 1 indicating the least deprived and 20 indicating the most deprived, was missing for `r imdtext`, because it is only recorded for English practices; alcohol status was missing for `r alcoholtext`; smoking status was missing for `r smokingtext`; BMI, or a calculated BMI from height and weight measurements, was missing for `r bmitext`, and baseline total cholesterol was missing for `r choltext`.

### Primary analysis

All results described below are from the fully adjusted model. The results of the age-only model are presented in Supplementary Figure 3, and demonstrate that adjustment for additional covariates had a minimal effect.

@@

As shown in Figure \@ref(fig:fig-results), our results show no effect of lipid-regulating agents on probable (`r probad_text`) and possible (`r possad_text`) Alzheimer's disease when compared with no treatment, with the exception of fibrates on probable Alzheimer's disease (`r probad_fib_text`).

@@

### Sensitivity analyses

When stratifying based on year of entry to the cohort, we observed no variation in risk by time period in any subgroup except for probable Alzheimer’s disease (Appendix XXX). Removing participants coded as female aged 55 and under at index from our analysis had minimal effect on our estimates (Appendix XXX).

<!-- For our negative control, there was some evidence that treatment with a lipid-regulating agent was associated with an increased risk of back pain (r sens_bp_text), suggesting there may be some residual confounding (Appendix XXX). -->

<!--- I actually think it would be better to include these results not in the appendix --->

<!-- By pregnancy -->

<!-- By age-at-entry -->

<!-- By statin type (lipo vs hydrophilic) -->

### Controls

<!-- Backpain -->

<!-- Coronary artery disease -->

_TBC - supervisors to email about it_

## Discussion

<!--- To what extent should I cross-reference with the other literature here? --->

### Main findings

@@

### Comparison to other literature

@@

Several previous studies have demonstrated that statins are the overwhelmingly favoured methods of cholesterol control. __[CITATION NEEDED]__

@@

These sources of evidence conflict with the findings of our analysis, where statin use was associated with an increased risk of all-cause dementia. However, some of the included studies in the meta-analysis specifically exclude vascular dementia from the definition of all-cause dementia,[@chao2015] which may lead to a artifical protective effect of statins on all-cause dementia

@@

Much less literature is available on the association between lipid-regulating agents and vascular dementia or other dementia. A recent review found four observational studies examining the association of statins and vascular dementia found no effect (RR:0.93, 95% CI 0.74–1.16).[@poly2020] This contrasts with the increased effect found in our analysis. An additional analysis found that lipophilic statins were more harmful than hydrophilic statins in vascular dementia, potentially due to their ability to cross the blood brain barrier.

@@

### Original contribution to research

There are two primary ways in which this research adds to the topic:

* _Size of the CPRD and length of follow-up:_ Having reviewed the other studies identified by the systematic review in Chapter 4, this analysis of 1.7 million participants is one of the largest studies of this research question.
* _Addressing the limitations of other observational analyses:_ Analyzing this data has provided the opportunity to use a separate analytical technique to many of the studies identified in the systematic review, As an example, the Hippsley-Cox BMJ paper examining the effect of statins, which makes use of the THIN EHR database, likely suffers from immortal time bias as exposed and unexposed participants are not followed up from a common time point.[@hippisley-cox2010] As touched on in the section above, this provides an additional evidence point with a different source and direction of bias, which is useful for the triangulation aspect of the thesis.

### Coding quality in the CPRD

Lots of things influence prevalence: historic trends, availability of good codes

There is some encouraging evidence that a dedicated coding exercise increases the identification of dementia case in primary care database, and does not add a substantial administrative burden.[@russell2013]

### Strengths and Limitations

@@

A further limitation is the possibility of time-varying confounding <!--- Either deal with this issue here, or keep the section above dealing with time-varying confounding. --->

@@

@@

### Enabling easy synthesis

In light of my own experiences in attempting to extract information for papers assessing preventative measures, as documented in Section \@ref(), the outputs from this analysis are described in detail in both. IN line with this, this analyses has been reported in line with the Strengthening the Reporting of Observational Studies in Epidemiology (STROBE) reporting guidelines[@vandenbroucke2007] (see Appendix \@ref() for the STROBE checklist) <!--- Maybe don't include the whole thing, or apply it to this thesis rather than the paper --->

Reporting guidelines do not always work, but help to guide accurate rep

All code and summary statistics (i.e. the tables presented in this chapter, plus summary tables of effect estimates) are readily available in machine readable formats (i.e. as comma separated values, or CSV, files) from the Github repository for this project. <!--- Will need to cite this, and may need to archive on Zenodo. --->

@@

### Conclusions

This chapter has provided new evidence on the potential repurposing of lipid-regulating agents for the prevention of all-cause dementia, Alzheimer’s disease, vascular dementia, and other dementia. It found use of lipid-regulating agents associated with a small reduction in probable or possible Alzheimer’s disease, and were associated with an increased risk of all-cause, vascular and other dementias. In all cases, the estimated associations were driven by those observed in the statin subgroup, which comprised the majority of participants in our cohort.

This chapter attempted to account for important sources of bias and provide a comparison with other available literature, as identified in the systematic review presented in Chapter \@ref(sys-rev-heading). However, there is a strong potential for unmeasured confounding, misclassification and reverse causation, which raises questions about our findings, in particular the unexpected increase in risk of vascular dementia associated with statin use.

Given the observed increased association of statins in other

While these

Future research should aim to address these potential biases and, while it may be costly in terms of time and resources, a large scale, long-term randomized controlled trial would provide useful additional information on the effect of lipid-regulating agents on the risk of dementia and related outcomes.

@@

## References

>>>>

---
title: 
date: "Maunscript last updated: `r format(Sys.time(), '%d %B, %Y')`"
bibliography: references.bib
csl: vancouver.csl
pandoc_args: [       "-F", "./pandoc-word-pagebreak"     ]
output:
  bookdown::word_document2:
      toc: false
      toc_depth: 3
      number_sections: false 
      reference_docx: "../../templates/word-styles-reference-01.docx"
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval = TRUE) # Remove to knit with plots/data
options(scipen = 999)
source(here::here("R/doc_options.R"))
source(here::here("R/helper.R"))

#Load relevant libraries and datasets from the 'outputs' folder

knitr::read_chunk(here::here("R/04-Code-CPRD-Analysis.R"))
doc_type <- knitr::opts_knit$get('rmarkdown.pandoc.to')

```

# Front matter

__Type of manuscript:__

Original Research Article

__Title: __

Association of lipid-regulating drugs with dementia and related conditions: an observational study of data from the CPRD

__Authors and affiliations:__

Luke A McGuinness^1,2,\*^, Julian PT Higgins^1,2^, Venexia M Walker^1,2^, Neil M Davies^1,2^, Richard M Martin^1,2^,  Elizabeth Coulthard^3^, George Davey-Smith^1,2^,  Patrick G Kehoe^3,4^ and Yoav Ben-Shlomo^2^

(1) MRC University of Bristol Integrative Epidemiology Unit, Bristol, UK
(2) Bristol Medical School: Population Health Sciences, University of Bristol, Bristol, UK
(3) Bristol Medical School: Translational Health Sciences, University of Bristol, Bristol, UK
(4) Dementia Research Group, Bristol Medical School, University of Bristol, Bristol, UK

__Corresponding author:__

Luke McGuinness; Bristol Medical School, University of Bristol,
Canynge Hall, 39 Whatley Road, Bristol, BS8 2PS, United Kingdom.; luke.mcguinness@bristol.ac.uk

__Conflict of interest statement__
None declared.

__Sources of funding__

LAM is supported by an NIHR Doctoral Research Fellowship (DRF-2018-11-ST2-048). JPTH and GDS are supported by NIHR Bristol Biomedical Research Centre at University Hospitals Bristol NHS Foundation Trust and the University of Bristol. YBS and JPTH are supported by the NIHR Applied Research Collaboration West (ARC West) at University Hospitals Bristol NHS Foundation Trust. LAM, VMW, NMD, RMM, JPTH and GDS are members of the MRC Integrative Epidemiology Unit at the University of Bristol. JPTH is a National Institute for Health Research (NIHR) Senior Investigator (NF-SI-0617-10145).
The views expressed in this article are those of the authors and do not necessarily represent those of the NHS, the NIHR, MRC, or the Department of Health and Social Care.

__Data/code availability__
This analysis used the CPRD-GOLD primary care dataset March 2016 snapshot (ISAC 15_246R), which is available upon application to the CPRD Independent Scientific Advisory Committee. The code lists used to define the outcomes and covariates for this study, in addition to the cleaning and analysis scripts used to create the study cohort and perform the analyses, are available on Github (https://github.com/mcguinlu/CPRD-LRA), and were archived at the time of submission on Zenodo (DOI: __TBC__).

@@

```{r attrition, echo = FALSE}
```

# Abstract

__Introduction:__ There is some evidence that circulating blood lipids play a role in the development of of Alzheimer's disease and dementia. As a result, these readily modifiable risk factors could be targeted by existing lipid-regulating agents, including statins, for the prevention of dementia. Here, we assess the association between lipid-regulating agents and subsequent risk of dementia and related conditions in over <!-- TODO replace -->
 participants.

__Methods:__ Cox proportional hazard models, allowing for a time-varying treatment indicator, were used to estimate the association between seven lipid-regulating drug classes and five dementia outcomes (all-cause, vascular and other dementia, and probable and possible Alzheimer's disease) in a prospective cohort study design using routinely collected data from the Clinical Practice Research Datalink.

__Results:__ We analyzed `r totaltext` with a average follow-up of `r fu_ab_text`. Lipid-regulating agents were found to have no effect on risk of Alzheimer's disease (probable AD `r probad_text`; possible AD `r possad_text`), but an increased risk of all-cause (`r anydem_text`), vascular (`r vasdem_text`) and other dementia (`r othdem_text`).

__Discussion:__ We found lipid-regulating agents had no effect on Alzheimer's disease risk, but increased the risk of all-cause, vascular and other dementia. However, this analysis is likely subject to residual confounding by indication and bias due to outcome misclassification.

__Keywords:__ Dementia; Alzheimer's disease; Lipids; Statins; Cohort study; Observational study

# Introduction

Dementia is a major neurocognitive disorder, the most common types of which are Alzheimer’s disease (AD) and vascular dementia (VaD).[@prince2016] Despite an increasing number of cases globally and decades of research, there remains much unknown about the pathogenesis and progression of the disease, and, at present, no effective treatment exists to arrest or reverse the cognitive decline associated with the condition.[@jobke2018] Drug repurposing, the identification of new applications for previously approved drugs, may provide an efficient mechanism to discover effective preventative and therapeutic treatments for dementia.[@ashburn2004; @pushpakom2019]

Several cardiovascular factors have been identified as potential risk factors for dementia,[@walker2019a] and of these, circulating lipid levels represent a promising target for intervention due to the ready availability of lipid-modifying treatments. In this context, determining whether lipid-regulating agents could be repurposed for the prevention of dementia and related diseases would be helpful in the development of evidence-based prevention policy. Several existing prospective studies have examined the association of lipid-regulating agent use with dementia.[@chu2018; @larsson2018; @ancelin2012; @bettermann2012;@rea2005] However, many of these studies are small, record few outcomes, and have limited follow-up.

The use of electronic health data for epidemiological research has several advantages.[@casey2016] As the data are collected through the routine care of a large cohort, they allow for nested prospective studies using sample sizes and time-scales which would be unfeasible using traditional methods. In addition, data are collected for care provision and without a specific research question in mind,  providing a holistic picture of a patient and their health experience. This provides vital data on a range of potential confounders which can be incorporated into an analysis.

We therefore aim to examine the association between several major classes of lipid-regulating agents and all-cause dementia, Alzheimer's disease, vascular dementia and other dementia, in the Clinical Practice Research Datalink (CPRD), a large, population-based electronic health record (EHR) database.[@herrett2015]

# Methods {#methods}

## Study design and protocol

@@

@@

## Study Cohort

Participants were included in our study cohort if their record contained any of the following index events: a Read code for a diagnosis of hypercholesterolemia or related condition; a Read code for prescription of a lipid-regulating agent (such as statins); a total cholesterol test result of >4mmol/L; or an LDL-c test result of >2mmol/L.

@@

## Exposures

@@

## Outcomes

We considered five outcomes as part of this analysis: probable  Alzheimer’s disease, possible Alzheimer’s disease, vascular dementia, other dementia, and a composite all-cause dementia outcome. When two or more outcomes were coded in a participant’s clinical record, a decision tree was used to differentiate between them (Supplementary Figure 1). The diagnosis date of the outcome was determined by the first record of a relevant code.

@@

## Covariates

The analysis was adjusted for a range of baseline covariates including sex, grouped year of entry into the cohort (<2000, 2000-2004, 2005-2009, >2010), Charlson co-morbidity index, Index of Multiple Deprivation (IMD), consultation rate, alcohol (current, former,  never), smoking (current, former,  never), BMI, baseline total cholesterol, and history of cardiovascular disease, coronary bypass surgery, coronary artery disease, peripheral arterial disease, hypertension, chronic kidney disease, and Type 1 and Type 2 diabetes. All
covariates were determined at index and definitions for each can be found in Supplementary Table 1.

@@

## Analysis plan

All analyses were performed in STATA 15. Cox proportional hazard models were used to estimate the hazard ratio and corresponding 95% confidence intervals, allowing for the potential for clustering by practice. To address the potential for immortal time bias, we employed a time-varying indicator of treatment status to correctly allocate time-at-risk to the exposed and unexposed groups.[@levesque2010] We ran two models, one unadjusted except for age, which was captured by using the participant's age as the time axis for the model,[@lamarca1998; @gail2009; @pencina2007] and a second adjusted for the additional covariates described above. Additional analyses stratified by outcome and drug class were also performed.

In the case of missing data, we used multiple imputation by chained equations (MICE) in STATA to create 20 imputed datasets. All covariates included in the analytic model were also included in the imputation model. The full imputation model is available for inspection (See [Data/Code availability](#data-code-avail) section).

@@

## Additional analyses

We performed a number of sensitivity analysis. We stratified the analysis by grouped year of entry to explore the potential for a time period effect. As statins are contraindicated in pregnancy,[@nationalinstituteforhealthandcareexcellence]  we ran the models described above but excluding participants coded as female below the age of 55. We also further stratified the statin exposure group into lipophilic (Atorvastatin, Lovastatin, Simvastatin, Cerivastatin) and hydrophilic (Pravastatin, Rosuvastatin, Fluvastatin) statins. Finally, we performed a negative control analysis, using the fully adjusted model to investigate the association between lipid-regulating agents and back pain.

@@

# Results {#results}

## Patient characteristics

@@

@@

## Missing data

Full covariate information was available for `r missingtext`. Five key variables had some missing data: IMD 2010 score, a proxy for socioeconomic position that is measured as twentiles with 1 indicating the least deprived and 20 indicating the most deprived, was missing for `r imdtext`, because it is only recorded for English practices; alcohol status was missing for `r alcoholtext`; smoking status was missing for `r smokingtext`; BMI, or a calculated BMI from height and weight measurements, was missing for `r bmitext`; baseline total cholesterol was missing for `r choltext`; and baseline LDL cholesterol was missing for `r ldltext`.

## Primary analysis

All results described below are from the fully adjusted model. The results of the model adjusted only for age and for age and sex are presented in Supplementary Figure 3. Adjustment for additional covariates beyond age and sex had a limited impact on the observed effect estimates, with the exception of the Probable AD outcome. In this case, adjustment for the full set of covariates attenuated to the null the protective effect observed when adjusting only for age and sex.

@@

As shown in Figure \@ref(fig:cprdPrimary), our results show no effect of lipid-regulating agents on probable (`r probad_text`) and possible (`r possad_text`) Alzheimer's disease when compared with no treatment, with the exception of fibrates on probable Alzheimer's disease (`r probad_fib_text`).

@@

## Sensitivity analyses

```{r sens-text, echo = FALSE}
```

When stratifying based on year of entry to the cohort, we observed no variation in risk by time period in any subgroup except for probable Alzheimer’s disease (Supplementary Figure 4). Removing participants coded as female aged 55 and under at index from our analysis had minimal effect on our estimates (Supplementary Figure 5).

For our negative control, there was some evidence that treatment with a lipid-regulating agent was associated with an increased risk of back pain (`r sens_bp_text`), suggesting there may be some residual confounding (Supplementary Figure 6).

# Discussion {#discussion}

## Main findings

<!--- Is this true - no effect? Is it not a small reduction rather than no effect? --->

@@

## Comparison to other literature

@@

@@

These sources of evidence conflict with the findings of our analysis, where statin use was associated with an increased risk of all-cause dementia (`r anydem_text`). However, some of the included studies in the meta-analysis specifically exclude vascular dementia from the definition of all-cause dementia,[@chao2015] which may lead to a artifical protective effect of statins on all-cause dementia

@@

Much less literature is available on the association between lipid-regulating agents and vascular dementia or other dementia. A recent review found four observational studies examining the association of statins and vascular dementia found no effect (RR:0.93, 95% CI 0.74–1.16).[@poly2020] This contrasts with the increased effect found in our analysis (`r vasdem_text`). An additional analysis found that lipophilic statins were more harmful than hydrophilic statins in vascular dementia, potentially due to their ability to cross the blood brain barrier.

@@

## Strengths and Limitations

@@

@@

# Conclusions {#conclusion}

We have provided new evidence on the potential repurposing of lipid-regulating agents for the prevention of all-cause dementia, Alzheimer’s disease, vascular dementia, and other dementia. We found use of lipid-regulating agents not associated with probable or possible Alzheimer’s disease, but were associated with an increased risk of all-cause, vascular and other dementias. In all cases, the estimated associations were driven by those observed in the statin subgroup, which comprised the majority of participants in our cohort.

We have attempted to account for important sources of bias in our analysis and provide a comparison with other available literature. However, there is a strong potential for unmeasured confounding, misclassification and reverse causation, which raises questions about our findings, in particular the unexpected increase in risk of vascular dementia associated with statin use. Future research should aim to address these potential biases and, while it may be costly in terms of time and resources, a large scale, long-term randomized controlled trial would provide useful additional information on the effect of lipid-regulating agents on the risk of dementia and related outcomes.

<!--------------------------------------------------------------------------->
(ref:cprdCharacteristics-caption) Patient characteristics by drug class. Summary statistics are presented as "% (N)" unless otherwise specified in the variable name.

```{r cprdCharacteristics-table, message=FALSE, results="asis", echo = FALSE}
```

@@

@@

<!--------------------------------------------------------------------------->
(ref:cprdPrimary-cap) Results from primary analyses of CPRD data using the fully adjusted model and participant age as the time scale.

```{r cprdPrimary, echo = FALSE, results="asis", fig.pos = "H", fig.cap='(ref:cprdPrimary-cap)', out.width='100%', fig.scap='(ref:cprdPrimary-scap)'}
knitr::include_graphics(here::here("figures","cprd-analysis","fp_p1.jpeg"))
```
<!--------------------------------------------------------------------------->

@@

# Bibliography
